Equalizer Programming Guide

### 2009-02-09 Developer/ProgrammingGuide/paper.tex
Programming Guide: Add section on eq::Object
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{15cm}
  {\Large\sffamily The official reference for creating parallel,
    scalable OpenGL\texttrademark\ applications with the Equalizer
    parallel rendering framework}\\\vspace{1cm}
  Version 1.5.3 for Equalizer 0.6+svn\\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.5.1   & Dec 08, 2008 & Minor updates based on the print version\\
    1.5.2   & Jan 10, 2009 & Added \sref{sCompositor}\\
    1.5.3   & Jan 13, 2009 & Added \sref{sNetwork}\\
    \hline
  \end{tabularx}}
\clearpage

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/executionFlow.pdf}
  {\caption{\small\label{fExecutionFlow}Parallel Rendering}}\vspace{-2ex}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/processes.pdf}
  {\caption{\small\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
The render client implements the rendering part of an application. Its
execution is passive, it has no main loop and is completely driven by
Equalizer, based on the rendering tasks received from the server. The
tasks are executed by calling the appropriate task methods (see
\sref{ssTaskMethods}) in the correct thread and context. The application
either implements the task methods with application-specific code or
uses the default methods provided by Equalizer.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/Pixel.pdf}
  {\caption{\label{fPixel}\small A pixel compound}}\vspace{-5ex}
\end{wrapfigure}
Pixel compounds are similar to 2D compounds. The frusta of the
source rendering units are modified so that each unit renders an evenly
distributed subset of pixels, as shown in \fig{fPixel}\footnote{3D model courtesy of AVS, USA.}

@@

\subsection{\label{ssTaskMethods}Task Methods}

@@

\subsubsection{\label{sFrustum}Frustum}
Compounds have a frustum description to define the physical layout of
the display environment.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
  {\caption{\small\label{fFrusta}Wall and projection parameters}}
\end{wrapfigure}
The frustum description is inherited by the children, therefore the
frustum is typically defined on the topmost compound.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Render Loop}}
\end{wrapfigure}
The model data is spatially organized in a 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree. During
this preprocessing step, each node of the tree gets a database range
assigned. The root node has the range [0, 1], its left child [0, 0.5]
and its right child [0.5, 1], and so on for all nodes in the tree. The
preprocessed model is saved in a binary format for accelerating
subsequent loading.

@@

The actual rendering uses display lists or vertex buffer objects. These
OpenGL objects are allocated using the object manager. The rendering is
done by the leaf nodes, which are small enough to store the vertex
indices in a \textsf{short} value for optimal performance with VBO's.
The leaf nodes reuse the objects stored in the object manager, or create
and set up new objects if it was not yet set up. Since one object
manager is used per thread (pipe), this allows a thread-safe sharing of
the compiled display lists or VBO's across all windows of a pipe.

@@

When using 2D (sort-first) or stereo decompositions, no special
programming is needed to achieve good scalability, as \textsf{eVolve} is
mostly fill-limited and therefore scales nicely in these modes.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\small\label{fSlices}Volume Rendering using 3D Textures}}
\end{wrapfigure}
For recomposition, the 2D frame buffer contents are blended to form a
seamless picture. For correct blending, the frames are ordered in the
same back-to-front order as the slices used for rendering, and use the
same blending parameters. Simplified, the frame buffer images are
`thick' slices which are `rendered' by writing their content to the
destination frame buffer using the correct order.

For orthographic rendering, determining the compositing order of the
input frames is trivial. The screen-space orientation of the volume
bricks determines the order in which they have to be composited. The
bricks in \textsf{eVolve} are created by slicing the volume along one
dimension. Therefore the range of the resulting frame buffer images,
together with the sorting order, is used to arrange the frames during
compositing. \fig{fBlendOrtho} shows this composition for one view.

\begin{wrapfigure}{l}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
Finding the correct assembly order for perspective frusta is more
complex. The perspective distortion invalidates a simple orientation
criteria like the one used for orthographic frusta. For the view and
frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
1-4-3-2.

@@

Statistics are enabled per entity using an attribute hint. The hint
determines how precise the gathered statistics are. When set to
\textsf{fastest}, the per-frame clock is sampled directly when the event
occurs. When set to \textsf{nicest}, all OpenGL commands will be
finished before sampling the event. This incurs a performance penalty,
but gives more correct results. The default setting is fastest in
release builds, and nicest in debug builds.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/statistics.png}
  {\caption{\small\label{fStatistics}Statistics for a two-node 2D compound}}
\end{wrapfigure}
The events are processed by the channel's and window's
\textsf{processEvent} method. The default implementation sends these
events to the config, using \textsf{Config::sendEvent}, as explained in
\sref{sEventHandling}. When the default implementation of
\textsf{Config::handleEvent} receives the statistics event, it sorts the
event per frame and per originator. When a frame has been finished, the
events are pushed to the local (app-)node for visualization.

@@

\subsection{\label{sNetwork}The Network Layer}

@@

\subsubsection{Connections}

@@

\subsubsection{Packet Handling}

@@

\subsubsection{Nodes}

@@

\subsubsection{Session}

@@

\subsubsection{Object}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/netObject.pdf}
  {\caption{\small\label{fNetObject}Communication between two objects}}
\end{wrapfigure}
Objects can send command packets to each other by subclassing
\textsf{eq::net::Ob\-ject\-Packet}, which is derived from
\textsf{SessionPacket}. The object packet extends the session packet by
the object identifier and an object instance identifier.

Multiple instances of the same object (identifier) can be mapped on the
same node and session simultaneously. Each object instance has a
node-unique instance identifier. When the instance ID is set to
\textsf{EQ\_ID\_ANY} (the default value), a command is dispatch to all
object instance registered to the the session with the correct object
ID. Multiple object instance per node are used when different threads on
the node need to process a different version of the same object.

@@

\subsubsection{Barrier}

@@

>>>>

\date{
  \vspace{15cm}
  {\Large\sffamily The official reference for creating parallel,
    scalable OpenGL\texttrademark\ applications with the Equalizer
    parallel rendering framework}\\\vspace{1cm}
  Version 1.5.4 for Equalizer 0.6+svn\\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
\listoffigures
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.5.1   & Dec 08, 2008 & Minor updates based on the print version\\
    1.5.2   & Jan 10, 2009 & Added \sref{sCompositor}\\
    1.5.3   & Jan 13, 2009 & Added \sref{sNetwork}\\
    1.5.4   & Feb 09, 2009 & Added \sref{sNetUsage}\\
    \hline
  \end{tabularx}}
\clearpage

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/executionFlow.pdf}
  {\caption{\small\label{fExecutionFlow}Parallel Rendering}}
  \vspace{-3ex}
\end{wrapfigure}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/processes.pdf}
  {\caption{\small\label{fProcesses}Equalizer Processes}}
\end{wrapfigure}
The render client implements the rendering part of an application. Its
execution is passive, it has no main loop and is completely driven by
Equalizer, based on the rendering tasks received from the server. The
tasks are executed by calling the appropriate task methods (see
\sref{sTaskMethods}) in the correct thread and context. The application
either implements the task methods with application-specific code or
uses the default methods provided by Equalizer.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-1ex}\includegraphics[width=.618\textwidth]{images/Pixel.pdf}
  {\caption{\label{fPixel}\small A pixel compound}}
  \vspace{-6ex}
\end{wrapfigure}
Pixel compounds are similar to 2D compounds. The frusta of the
source rendering units are modified so that each unit renders an evenly
distributed subset of pixels, as shown in \fig{fPixel}\footnote{3D model courtesy of AVS, USA.}

@@

\subsection{\label{sTaskMethods}Task Methods}

@@

\subsubsection{\label{sFrustum}Frustum}
\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frusta.pdf}
  {\caption{\small\label{fFrusta}Wall and projection parameters}}
  \vspace{-3ex}
\end{wrapfigure}
Compounds have a frustum description to define the physical layout of
the display environment.

The frustum description is inherited by the children, therefore the
frustum is typically defined on the topmost compound.

@@

The model data is spatially organized in a 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree. During
this preprocessing step, each node of the tree gets a database range
assigned. The root node has the range [0, 1], its left child [0, 0.5]
and its right child [0.5, 1], and so on for all nodes in the tree. The
preprocessed model is saved in a binary format for accelerating
subsequent loading.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \vspace{-2ex}\includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Render Loop}}
\end{wrapfigure}
The actual rendering uses display lists or vertex buffer objects. These
OpenGL objects are allocated using the object manager. The rendering is
done by the leaf nodes, which are small enough to store the vertex
indices in a \textsf{short} value for optimal performance with VBO's.
The leaf nodes reuse the objects stored in the object manager, or create
and set up new objects if it was not yet set up. Since one object
manager is used per thread (pipe), this allows a thread-safe sharing of
the compiled display lists or VBO's across all windows of a pipe.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/slices.pdf}
  {\caption{\small\label{fSlices}Volume Rendering using 3D Textures}}
\end{wrapfigure}
When using 2D (sort-first) or stereo decompositions, no special
programming is needed to achieve good scalability, as \textsf{eVolve} is
mostly fill-limited and therefore scales nicely in these modes.

@@

For recomposition, the 2D frame buffer contents are blended to form a
seamless picture. For correct blending, the frames are ordered in the
same back-to-front order as the slices used for rendering, and use the
same blending parameters. Simplified, the frame buffer images are
`thick' slices which are `rendered' by writing their content to the
destination frame buffer using the correct order.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
For orthographic rendering, determining the compositing order of the
input frames is trivial. The screen-space orientation of the volume
bricks determines the order in which they have to be composited. The
bricks in \textsf{eVolve} are created by slicing the volume along one
dimension. Therefore the range of the resulting frame buffer images,
together with the sorting order, is used to arrange the frames during
compositing. \fig{fBlendOrtho} shows this composition for one view.

Finding the correct assembly order for perspective frusta is more
complex. The perspective distortion invalidates a simple orientation
criteria like the one used for orthographic frusta. For the view and
frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
1-4-3-2.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/statistics.png}
  {\caption{\small\label{fStatistics}Statistics for a two-node 2D compound}}
\end{wrapfigure}
Statistics are enabled per entity using an attribute hint. The hint
determines how precise the gathered statistics are. When set to
\textsf{fastest}, the per-frame clock is sampled directly when the event
occurs. When set to \textsf{nicest}, all OpenGL commands will be
finished before sampling the event. This incurs a performance penalty,
but gives more correct results. The default setting is fastest in
release builds, and nicest in debug builds.

The events are processed by the channel's and window's
\textsf{processEvent} method. The default implementation sends these
events to the config, using \textsf{Config::sendEvent}, as explained in
\sref{sEventHandling}. When the default implementation of
\textsf{Config::handleEvent} receives the statistics event, it sorts the
event per frame and per originator. When a frame has been finished, the
events are pushed to the local (app-)node for visualization.

@@

\section{\label{sNetwork}The Network Layer}

@@

\subsection{Connections}

@@

\subsection{Packet Handling}

@@

\subsection{Nodes}

@@

\subsection{Session}

@@

\subsection{\label{sNetObject2}Object}

@@

Objects can send command packets to each other by subclassing
\textsf{eq::net::Ob\-ject\-Packet}, which is derived from
\textsf{SessionPacket}. The object packet extends the session packet by
the object identifier and an object instance identifier.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/netObject.pdf}
  {\caption{\small\label{fNetObject}Communication between two objects}}
\end{wrapfigure}
Multiple instances of the same object (identifier) can be mapped on the
same node and session simultaneously. Each object instance has a
node-unique instance identifier. When the instance ID is set to
\textsf{EQ\_ID\_ANY} (the default value), a command is dispatch to all
object instance registered to the the session with the correct object
ID. Multiple object instance per node are used when different threads on
the node need to process a different version of the same object.

@@

\subsection{Barrier}

@@

\subsection{\label{sNetUsage}Usage in Equalizer}

The Equalizer client library and server are built on top of the network
layer. They influence the design of it, and serve as a sample
implementation on how to use classes in the \textsf{eq::net} namespace.

\subsubsection{eq::Object}

Equalizer uses the \textsf{eq::net::Object} data distribution mechanism
for implementing data distribution between the server, application and
render clients. To do this, \textsf{eq::Object} implements one usage
pattern for the \textsf{eq::net::Object}, allowing inheritance with data
distribution. It is easier to use, but imposes one typical way to implement data
distribution.

The \textsf{eq::Object} data distribution is centered around the concept
of dirty bits. Dirty bits form a 64-bit mask which marks which parts of
the object have to be distributed during the next commit.

For serialization, the default \textsf{eq::net::Object} serialization
functions are implemented by \textsf{eq::Object}, which (de-)serializes
and resets the dirty bits, and calls \textsf{serialize} or
\textsf{deserialize} with the bit mask specifiying which data has to be
transmitted or received. During a commit or sync, the current dirty bits
are given, whereas during object mapping all dirty bits are passed to
the serialization methods.

In order to use \textsf{eq::Object}, the following steps have to be
taken:

\begin{description}
\item[Inherit from \textsf{eq::Object}:] The base class will provide the
  dirty bit management and call serialize and deserialize
  appropriately. By overriding \textsf{getChangeType}, the default
  versioning strategy might be changed.
\item[Define new dirty bits:] Define dirty bits for each data item by
  starting at \textsf{Object::DIRTY\_CUSTOM}, shifting this value
  consecutively by each new dirty bit.
\item[Implement serialize and deserialize:] For each object-specific
  dirty bit which is set, stream the corresponding data item to or from
  the provided stream. Call the parent method first in both
  functions. For application-specific objects, write a (de-)serialization
  function.
\item[Mark dirty data:] In each 'setter' method, call \textsf{setDirty}
  with the corresponding dirty bit.
\end{description}

The inheritance chain \textsf{eq::View} $\rightarrow$
\textsf{eq::Frustum} $\rightarrow$ \textsf{eq::Object} $\rightarrow$
\textsf{eq::net::Object} serves as sample implementation for this
process. The \textsf{eq::Object} base class provides the facility to set
a name on the object, which is automatically distributed.

The registration and mapping of \textsf{eq::Object}s is done in the same
way as for \textsf{eq::net::Object}s, which has been described in
\sref{sNetObject2}. Equalizer objects derived from \textsf{eq::Object},
e.g., \textsf{eq::View}, are typically registered and mapped by
Equalizer.
